---
output: 
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# snipR

snipR implements the blocked sorted neighbors algorithm for discovering potential duplicate pairs. The analysis is broken down into a series of steps:

* Initialize Neighbors object

* Perform BSN and collect possible duplicate pairs

* Score duplicate pairs and retain those above specified threshold

* Export, repeat, or verify accuracy of algorithm

## Installation

You can install bsnR from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("mPloenzke/snipR")
```

If you encounter a bug, please file a minimal reproducible example on [github](https://github.com/mPloenzke/snipR/issues). 

## Learning snipR

To get started, please read the intro vignette: `vignette("Introduction", package = "snipR")`. 

## Key data structures

The base S3 object in bsnR is a _Neighbors_, a contained for the raw data and user specified keys. At each step of the analysis this object will be promoted, allowing subsequent steps to be performed. 
